Numerical and asymptotic solution of nonlinear electrokinetic phenomena

### 2012-03-27 notes/article/main.tex
Fix figures.
Work on paper.

The paper is structured as follows: the governing equations of the electrokinetic model
are derived in Section \ref{sec:equations}. The algorithm for the discretization and 
the solution of the system is described in Section \ref{sec:algorithm}.
The non-linear analytical asymptotic solution are presented in Section{sec:asymp}.
The numerical results and their comparison to the analytical asymptotic solution 
are presented and discussed in Section \ref{sec:results}.

@@

We consider a spherical ion-exchanger. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we assume zero slip, anion impermeability, cation selectivity 
(with fast Butler-Volmer\cite{bard2000book} kinetics
$\bn \cdot \bj_+ = k (1 - c_+/\gamma)$, where $k \rightarrow \infty$),
and high conductance of the particle (yielding a uniform electric potential):
\begin{eqnarray*}
\bv & = & \bzero, \\
\bn \cdot \bj_- & = & 0, \\
c_+ &=& \gamma, \\
\varphi & = & \cV.
\end{eqnarray*}

@@

\subsubsection{Bulk-scale effective boundary conditions}
By integrating the microscale equations across the Debye layer
(see the full derivation in \cite{yariv2010asymptotic}), 
the effective boundary conditions at the ion-exchanger surface are:
\begin{eqnarray*}
0 &=& \varPhi + \log C, \\
0 &=& \bnhat \cdot \bnabla \pars{\varPhi - \log C}, \\
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C.
\end{eqnarray*}
where $\zeta = \cV - \varPhi = \log (C / \gamma)$ 
is the zeta potential (the voltage drop across the Debye layer).

@@

The iterative solver above can be used in principle to compute the steady-state solution for any given $\beta$.
For $\beta \ll 1$, the linear terms are the dominant ones, 
so the solution is approximately linear in $\beta$ 
(as derived in \cite{yariv2010migration}), and the solver convergence is fast.

@@

In order to validate the numerical solver, an asymptotic analysis of the nonlinear system
is performed as described in \ref{app:asymp}. The numerical results are compared 
to the asymptotic solutions, and good correspondence is achieved.

This phenomenon is confirmed both by the numerical
results and by asymptotic expansion of the nonlinear system (for $\alpha=0$).
The switch from the linear regime to the cubic regime occurs at a critical electric
field value
\begin{eqnarray*}
\beta_c = O(\sqrt{|\gamma - 1|}).
\end{eqnarray*}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=2.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=5.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=10.00.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=100.00.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma > 1$. 
        The analytic solution velocity is represented by dashed lines, and the numerical
        results are represented by squares.}
    \label{fig:Linear1}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.50.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.20.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.10.eps}
    \includegraphics[width=0.45\textwidth]{figs/Linear/g=0.01.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by solid lines, and the numerical
        results are represented by circles.}
    \label{fig:Linear2}
    \end{center}
\end{figure}

\begin{appendix}
\section{Asymptotic Analysis} \label{app:asymp}
The nonlinear system is defined by the following PDEs:
\label{PDEs}
\begin{eqnarray*}
\bnabla \cdot \pars{C \bnabla \varPhi} &=& 0, \\
\bnabla \cdot \pars{\bnabla C - \alpha \bV C} &=& 0, \\
\bnabla \cdot \bV &=& 0, \\
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi &=& \bzero.
\end{eqnarray*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear Solution $(k=1)$} \label{app:linear}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nonlinear Solution ($k > 1$)} \label{app:cubic}
Note that, due to symmetry considerations, $\cU(\beta)$ is an anti-symmetric function:
\begin{eqnarray*}
\cU(-\beta) &=& -\cU(\beta).
\end{eqnarray*}
Therefore, $\cU_2 = 0$ and the next velocity term is the cubic one:
\begin{eqnarray*}
\cU(\beta) &\approx& \beta \cU_1 + \beta^3 \cU_3 + O(\beta^5).
\end{eqnarray*}

@@

\end{appendix}

>>>>

The paper is structured as follows: the governing equations of the electrokinetic model
are derived in Section \ref{sec:equations}. The algorithm for the discretization and 
the solution of the system is described in Section \ref{sec:algorithm}.
The analytical asymptotic solution for the non-linear system is derived 
and presented in Section \ref{sec:asymp}.
The numerical results and their comparison to the analytical asymptotic solution 
are presented and discussed in Section \ref{sec:results}.

@@

We consider a spherical ion-exchanger. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we assume high conductance of the particle (yielding a uniform electric potential),
zero slip, anion impermeability, cation selectivity 
(with fast Butler-Volmer\cite{bard2000book} kinetics
$\bn \cdot \bj_+ = k (1 - c_+/\gamma)$, where $k \rightarrow \infty$):
\begin{eqnarray*}
\varphi & = & \cV, \\
\bv & = & \bzero, \\
\bn \cdot \bj_- & = & 0, \\
c_+ &=& \gamma, \\
\end{eqnarray*}
where $\gamma$ is the ratio between the cationic concentration on the surface and the cationic
concentration in the fluid bulk.

@@

\subsubsection{Bulk-scale effective boundary conditions}
By integrating the microscale equations across the Debye layer
(see the full derivation in \cite{yariv2010asymptotic}), 
the effective boundary conditions at the ion-exchanger surface are:
\begin{eqnarray*}
0 &=& \varPhi + \log C, \\
0 &=& \bnhat \cdot \bnabla \pars{\varPhi - \log C}, \\
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C,
\end{eqnarray*}
where $\zeta = \cV - \varPhi = \log (C / \gamma)$ 
is the zeta potential (the voltage drop across the Debye layer).

@@

The iterative solver above can in principle be used 
to compute the steady-state solution for any given $\beta$.
For $\beta \ll 1$, the linear terms are the dominant ones, 
so the solution is approximately linear in $\beta$ 
(as derived in \cite{yariv2010migration}), and the solver convergence is fast.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Asymptotic Analysis} \label{sec:asymp}
The nonlinear system is defined by the following PDEs:
\label{PDEs}
\begin{eqnarray*}
\bnabla \cdot \pars{C \bnabla \varPhi} &=& 0, \\
\bnabla \cdot \pars{\bnabla C - \alpha \bV C} &=& 0, \\
\bnabla \cdot \bV &=& 0, \\
\bLaplacian \bV - \bnabla P + \Laplacian \varPhi \bnabla \varPhi &=& \bzero.
\end{eqnarray*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linear Solution $(k=1)$} \label{app:linear}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Nonlinear Solution ($k > 1$)} \label{app:cubic}
Note that, due to symmetry considerations, $\cU(\beta)$ is an anti-symmetric function:
\begin{eqnarray*}
\cU(-\beta) &=& -\cU(\beta).
\end{eqnarray*}
Therefore, $\cU_2 = 0$ and the next velocity term is the cubic one:
\begin{eqnarray*}
\cU(\beta) &\approx& \beta \cU_1 + \beta^3 \cU_3 + O(\beta^5).
\end{eqnarray*}

@@

In order to validate the numerical solver, an asymptotic analysis of the nonlinear system
is performed as described in \ref{sec:asymp}. The numerical results are compared 
to the asymptotic solutions, and good correspondence is achieved.

This phenomenon is confirmed both by the numerical
results and by asymptotic expansion of the nonlinear system (for $\alpha=0$).
The switch from the linear regime to the cubic regime occurs at a critical electric
field value $\beta_c = O(\sqrt{|\gamma - 1|})$.

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.8\textwidth]{figs/linearNeg.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma > 1$. 
        The analytic solution velocity is represented by dashed lines, and the numerical
        results are represented by squares.}
    \label{fig:Linear1}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
    \includegraphics[width=0.8\textwidth]{figs/linearPos.eps}
    \caption{Steady-state velocity $\cU$ as a function of the 
        electric field magnitude $\beta$, for $\gamma < 1$. 
        The analytic solution velocity is represented by solid lines, and the numerical
        results are represented by circles.}
    \label{fig:Linear2}
    \end{center}
\end{figure}
